---
title: "Incremental CEA Basics"
author: "Justin Ingels"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ceaR Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The package ceaR is designed to facilitate cost-effectiveness analyses of 
individual-level data. In most cases, the interested user will have a dataset
of individuals that fall into one of two (or more) treatment or intervention
conditions. At a minimum, each individual should have a cost, effect, and an
indicator of the intervention or condition of which he or she is a member.
There may also be one or more covariate values for each individual including
baseline costs or effects, demographic, environmental, or disease-specific 
factors.

## Setup
The cea_setup function is used to define the variables representing costs, 
effects, and intervention condition as well as any covariate variables that
should be included in analyses and whether these covariates are applied to 
both the cost and effect analyses or just one or the other. Further, an
optional setting eff_more_better can be set to False in cases where more of
the effect is not a better outcome for an individual. An example would be if
the effect is a binary variable where a 1 represents death or disease.

As an example, the dataset clintrial_cea will be used... The formula method can
be used where the baseline costs (blcost) and QALYs (blqaly) will be covariates
with cost and qaly, respectively.

```{r}
library(ceaR)
testmodel <- cea_setup(cost | qaly ~ blcost | blqaly, intv = "treat", 
                       clintrial_cea)
```

The object testmodel is of class ceamodel and includes a data frame including
only the variables from clintrial_cea including in the function call. There are
also references to the cost, effect, intervention, and covariate variables; cost
and effect formulas, and the number of individuals in the dataset.

## Incremental CEA
To conduct an incremental cost-effectiveness analysis, the next step is to pass
the testmodel object to the ceamodel_incremental function. This function will
estimate delta_e and delta_c between conditions. The analyses are conducted
based on a regression model for both costs and effects. When covariates are
not included in the model, the results are equivalent to taking the differences
in the mean costs (and effects) by condition across individuals in the sample.
The results of this analysis are attached to the object of class ceamodel that
is passed to the function.

```{r}
testmodel <- ceamodel_incremental(testmodel)
```

